Hydra Battles and Cie

### 2021-04-13 doc/chapter-primrec.tex
small changes

\section{Introduction}
The definition of primitive recursive functions we use comes
from Russel O'Connors formalization in \coq{} of
G\"odel's incompleteness theorems~\cite{OConnor05}, now hosted in
the \texttt{theories/ordinals/Ackermann} directory.

@@

The traditional definition of the set of primitive recursive functions is structured as an inductive definition 
in five rules: three base cases, and two construction rules. Primitive recursive functions are total functions from $\mathbb{N}^n$ to
$\mathbb{N}$, for some $n\in\mathbb{N}$.

@@

\subsection{A little bit of semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$ , together with the  function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

Ok, the term \texttt{bigPR} looks to be a primitive recursive definition of the factorial function, although we haven't proved this fact yet. Fortunately, we will see in the following sections much simpler ways to prove that a given function is primitive recursive, whithout looking at an unreadable term.

@@

\begin{remark}
In the library \texttt{primRec}, all these lemmas are opaque (registered with \texttt{Qed}. Thus they do not allow the user to look at the witness of a proof of a \texttt{isPR} statement. Our example of page\pageref{sect:bigfac} was built using a  copy of \texttt{primRec.v} where many \texttt{Qed}s have been replaced with
\texttt{Defined}s.
\end{remark}

@@

Thus, our first subgoal is solved easily and the proof ends, applying already proven lemmas.

@@

\index{primrec}{Exercises}
\begin{exercise}
There is a lot of lemmas similar to \texttt{filter010IsPR} in the \texttt{primRec} library, useful to control the arity of functions.
Thus, the reader may look at them, and invent simple examples of application for each one.
\end{exercise}

@@

The following proof decomposes the \texttt{double} function as the composition of 
multiplication with the identity and the constant function wihich returns $2$.
\emph{Note that the lemma \texttt{const1\_NIsPR} considers this function as an unary function (unlike \texttt{const0\_NIsPR})}.

@@

The three firts lemmas make us sure that our function 
\texttt{Ack} satifies the ``usual'' equations.

@@

\index{coq}{Dependently typed functions}
\begin{Coqsrc}
Fixpoint max_v {n:nat} : forall (v: Vector.t nat n) , nat :=
  match n as n0 return (Vector.t nat n0 -> nat)
  with
    0 => fun v => 0
  | S p => fun (v : Vector.t nat (S p)) =>
             (max (Vector.hd v) (max_v  (Vector.tl v)))
  end.

>>>>

\section{Introduction}
Primitive recursive functions are a small class of total functions, corresponding to the expressive power of a simple imperative programming language without \textbf{while} loops, in which every program execution terminates.

Primitive recursive functions are total functions from $\mathbb{N}^n$ to
$\mathbb{N}$, for some $n\in\mathbb{N}$. Note that not all 
total $n$-ary recursive functions are primitive recursive
(see for instance Sect.~\vref{sect:ack-not-PR}).

The traditional definition of the set of primitive recursive functions is structured as an inductive definition 
in five rules: three base cases, and two recursive construction rules.

Please note the use of dots: $\ldots$ in the definition above. 
Dots are not part of \gallina's syntax. Thus, the formal definition of the set of primitive recursive function will have to overcome this representation problem.

  We present in this chapter a formalization of  primitive recursive functions, taken from  Russel O'Connor's formalization in \coq{} of
G\"odel's incompleteness theorems~\cite{OConnor05}.

\begin{remark}
 The theory of primitive recursive function is now hosted in
the \texttt{theories/ordinals/Ackermann} directory.
The specific part on G\"odel's theorem,  is also on
coq-community (\url{https://github.com/coq-community/goedel}) and requires the 
\href{https://github.com/coq-community/pocklington}{Pocklington library} for lemmas on primality.
\end{remark}

@@

@@

\subsection{A little bit of semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$.
This function is definied by mutual recursion,  together with the  function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

After this test, the term \texttt{bigPR} looks to be a primitive recursive definition of the factorial function, although we haven't proved this fact yet. Fortunately, we will see in the following sections simple ways to prove that a given function is primitive recursive, whithout building such an unreadable term.

@@

\begin{remark}
In the library \texttt{primRec}, all these lemmas are opaque (registered with \texttt{Qed}. Thus they do not allow the user to look at the witness of a proof of a \texttt{isPR} statement. Our example of page\pageref{sect:bigfac} was built using a  copy of \texttt{primRec.v} where many \texttt{Qed}s have been replaced with
\texttt{Defined}s.

If it does not cause compatibility problems (with 
\href{https://github.com/coq-community/goedel}{goedel library} fr instance), we plan to make all theses lemmas transparent.
\end{remark}

@@

Thus, our first subgoal is solved easily. The rest of the proof 
is just an application of already proven lemmas.

@@

\index{primrec}{Exercises}
\begin{exercise}
There is a lot of lemmas similar to \texttt{filter010IsPR} in the \texttt{primRec} library, useful to control the arity of functions.
Thus, the reader may look at them, and invent simple examples of application for each lemma.
\end{exercise}

@@

The following proof decomposes the \texttt{double} function as the composition of 
multiplication with the identity and the constant function which returns $2$.
\emph{Note that the lemma \texttt{const1\_NIsPR} considers this function as an unary function (unlike \texttt{const0\_NIsPR})}.

@@

The three first lemmas make us sure that our function 
\texttt{Ack} satifies the ``usual'' equations.

@@

\index{coq}{Dependently typed functions}
\begin{Coqsrc}
Fixpoint max_v {n:nat} : forall (v: Vector.t nat n) , nat :=
  match n as n0 return (Vector.t nat n0 -> nat)
  with
    0 => fun v => 0
  | S p => fun (v : Vector.t nat (S p)) =>
             max (Vector.hd v) (max_v  (Vector.tl v))
  end.
